### 2021-04-12 paper.Rmd
use American spelling

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings are revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualise cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

All of the aforementioned techniques can be enhanced through the use of interactivity.
By combining views with interactive elements like tooltips or highlighting the analysts is able to quickly interrogate interesting features of the data. One  particularly important interaction technique in the history of statistical graphics is called brushing [@Becker1987-dw]. When brushing an analyst drags their mouse over the view which results in a region being drawn onto the canvas. When there are multiple views present the act of brushing can be thought of as a database query; points that fall inside the brush can be used as a lookup to highlight or filter data on adjacent views (Figure \ref{fig:brushing}). This technique is particularly useful when combined with the tour (Section \@ref(sec:spin)).

@@

Below introduces the notations used for describing $d$-dimensional projections in a $p$-dimensional space. Let $X_{n \times p}$ be the data whose projections are of interest. A projection basis, $A_{p \times d}$, characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires each column to be a unit vector and to be perpendicular to each other. This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of the 2D palmerpenguins data (bill measures only) in a huber's plot, along with a 1D projection in the histogram. This projection is created based on projecting the data onto the solid line in the huber's plot and the it shows the separation among three species.

@@

Instead of randomly generating targets, guided tour [@Cook1995-ae] finds its targets with a technique called projection pursuit [@Friedman1974-ck].  In projection pursuit, an index function is a surjective mapping from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. Guided tour iteratively optimises the index function in the projection pursuit and produces the most interesting projection according to the index measure. The right panel of Figure \ref{fig:notation-target-gen} shows an example of a guided tour path.

@@

Physics models often consider $\mathcal{O}(10)$ free parameters and are compared to a much larger number of experimental observables ($\mathcal{O}(100)$). The comparison typically relies on numeric computation of the model predictions for all observables, potentially obscuring the nature of their dependence on the model parameters. Here, multivariate visualisation methods can provide new insights. This was demonstrated in @Cook2018-jm. For example, we explored grouping of experiments based on how they constrain the parameter space and we also identified an interesting multivariate outlier that pointed to potential issues with the data point. For illustration we show static views from the tour showing the orthogonal structure of the three groups (indicated by color), and a display which highlights the outlyingness of the data point marked with an asterisk symbol in Figure \ref{fig:physics}.

>>>>

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings are revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualize cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

All of the aforementioned techniques can be enhanced through the use of interactivity.
By combining views with interactive elements like tool tips or highlighting the analysts is able to quickly interrogate interesting features of the data. One  particularly important interaction technique in the history of statistical graphics is called brushing [@Becker1987-dw]. When brushing an analyst drags their mouse over the view which results in a region being drawn onto the canvas. When there are multiple views present the act of brushing can be thought of as a database query; points that fall inside the brush can be used as a look-up query to highlight or filter data on adjacent views (Figure \ref{fig:brushing}). This technique is particularly useful when combined with the tour (Section \@ref(sec:spin)).

@@

Below introduces the notations used for describing $d$-dimensional projections in a $p$-dimensional space. Let $X_{n \times p}$ be the data whose projections are of interest. A projection basis, $A_{p \times d}$, characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires each column to be a unit vector and to be perpendicular to each other. This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of the 2D palmerpenguins data (bill measures only) in a Huber plot, along with a 1D projection in the histogram. This projection is created based on projecting the data onto the solid line in the Huber plot and the it shows the separation among three species.

@@

Instead of randomly generating targets, guided tour [@Cook1995-ae] finds its targets with a technique called projection pursuit [@Friedman1974-ck].  In projection pursuit, an index function is a surjective mapping from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. Guided tour iteratively optimizes the index function in the projection pursuit and produces the most interesting projection according to the index measure. The right panel of Figure \ref{fig:notation-target-gen} shows an example of a guided tour path.

@@

Physics models often consider $\mathcal{O}(10)$ free parameters and are compared to a much larger number of experimental observables ($\mathcal{O}(100)$). The comparison typically relies on numeric computation of the model predictions for all observables, potentially obscuring the nature of their dependence on the model parameters. Here, multivariate visualisation methods can provide new insights. This was demonstrated in @Cook2018-jm. For example, we explored grouping of experiments based on how they constrain the parameter space and we also identified an interesting multivariate outlier that pointed to potential issues with the data point. For illustration we show static views from the tour showing the orthogonal structure of the three groups (indicated by color), and a display which highlights the "outlyingness" of the data point marked with an asterisk symbol in Figure \ref{fig:physics}.
